---
title: "utah_avy_dash"
format: dashboard
editor: visual
css: stylez.css
orientation: columns
---

```{r setup}
#| message: FALSE

library(knitr)
library(plotly)
library(tidyverse)
library(dataRetrieval)
library(lubridate)
library(sf)
library(ggspatial)
library(prettymapr)
library(lubridate)
library(rosm)
library(bsicons)
library(viridis)
library(shiny)
library(leaflet)
library(reactable)
library(ggridges)
```

```{r}
#| warning: FALSE
#| message: FALSE

# read csv
avalanches <- read.csv("avalanches.csv")


# Read zones
zones <- st_read("avalanche_center_forecast_zones.shp", quiet = TRUE)

zones <- st_transform(zones, 4326)


# cleaning names and new df
avalanches <- avalanches |>
  rename(Who = Trigger,
         Trigger = Trigger..additional.info) 

utah_avy <- avalanches |>
  select(-Comments.4, - Comments.3, -Comments.2, - Comments.1, - Weather.Conditions.and.History, -Terrain.Summary, -Accident.and.Rescue.Summary)

# creating season DF
utah_avy$Date <- as.Date(utah_avy$Date, format = "%m/%d/%Y")

utah_avy <- utah_avy |>
  filter(Date >= as.Date("2020-01-01") &
         Date <= as.Date("2025-07-01")) |>
  filter(Region %in% c("Abajos", "Logan", "Moab", "Ogden", "Provo", "Salt Lake", "Skyline", "Southwest", "Uintas"))



# split into lat and long
utah_avy <- utah_avy |>
  separate(Coordinates, into = c("latitude", "longitude"), sep = ",") |>
  mutate(
    latitude = as.numeric(latitude),
    longitude = as.numeric(longitude)
  )

utah_avy <- utah_avy |>
  filter(!is.na(longitude) & !is.na(latitude))

# convert to spatial points
utah_avy <- st_as_sf(utah_avy,
                       coords = c("longitude", "latitude"),
                       crs = 4326)

# Match Shapefile CRS

utah_avy <- st_transform(utah_avy, st_crs(zones))


# Depth 

utah_avy <- utah_avy |>
  mutate(
    Depth = case_when(
      str_detect(Depth, "'") ~ as.numeric(str_remove(Depth, "'")) * 12,
      str_detect(Depth, '"') ~ as.numeric(str_remove(Depth, '"')),
      TRUE ~ as.numeric(Depth)
    )
  )  |>
  mutate(
    Depth = Depth / 12
  )

# Width

utah_avy <- utah_avy |>
  mutate(
    Width = case_when(
      str_detect(Width, "'") ~ as.numeric(str_remove(Width, "'")),
      TRUE ~ as.numeric(Width)
    )
  )

# Vertical

utah_avy <- utah_avy |>
  mutate(
    Vertical = Vertical |>
      str_replace_all(",", "") |>
      str_replace_all("'", "") |>
      as.numeric()
  )

# Aspect

utah_avy$Aspect <- as.factor(utah_avy$Aspect)

# Create year column

utah_avy <- utah_avy |>
  mutate(year = format(Date, "%Y"),
         month = month(Date, label = TRUE, abbr = TRUE))

# Create size Column

utah_avy <- utah_avy |>
  mutate(
    size = Width * Vertical * Depth)
  



```

```{r}
#| message: FALSE
#| warning: FALSE
# Map

region_pal <- colorFactor(
  palette = "Set1",
  domain = zones$name
)


region_leaf <- leaflet() |>
addProviderTiles("Esri.WorldTopoMap") |>
  addPolygons(
    data = zones,
    color = ~region_pal(name),
    fillColor =  ~region_pal(name),
    fillOpacity = .6
    ) |>
  addCircleMarkers(
    data = utah_avy,
    color = ~region_pal(Region),
    fillColor = ~region_pal(Region),
    fillOpacity = 1,
    radius = 6,
    stroke = F
  ) |>
  addLegend(
    "bottomright",
    pal = region_pal,
    values = utah_avy$Region,
    title = "Region"
  )

```

```{r}
#| message: FALSE
#| warning: FALSE

avy_counts <- utah_avy |>
  st_drop_geometry() |>
  filter(
    !is.na(year),
    !is.na(Region), Region != "",
    !is.na(Weak.Layer), Weak.Layer != ""
  ) |>
  count(year, Region, Weak.Layer)

year_slide <- plot_ly(
  data = avy_counts,
  x = ~Region,
  y = ~n,
  frame = ~year,
  type = "bar",
  color = ~Weak.Layer
) |>
  layout(barmode = "stack") 
```

```{r}
#| message: FALSE
#| warning: FALSE
#| fig-format: png
#| fig-bg: transparent

# Make Facet DF
facets <- utah_avy |>
  filter(`Weak.Layer` == "Facets") |>
  select(Width, Depth, Aspect, Vertical, year, size, month) |>
  filter(
    !is.na(year),
    !is.na(Aspect), Aspect != "",
    !is.na(Width), Width != "",
    !is.na(Depth), Depth != "",
    !is.na(size), size != "",
    !is.na(month), month != "",
  )

# Mean facet depth by aspect
facet_average <- facets |>
  group_by(Aspect) |>
  summarise(
    count = n(),
    mean_depth = mean(Depth, na.rm = TRUE),
    mean_width = mean(Width, na.rm = TRUE), 
    mean_vertical = mean(Vertical, na.rm = TRUE),
    mean_size = mean(size, na.rm = TRUE)
  ) |>
  mutate(
    mean_depth = round(mean_depth, 1),
    mean_width = round(mean_width, 1),
    mean_vertical = round(mean_vertical, 1),
    mean_size = round(mean_size, 1)
  )
 

facet_aspect <-
  ggplot(facets, aes(x = Aspect, y = Depth, color = Aspect)) +
  geom_point() +
  geom_violin(alpha = 0.6)+
  geom_boxplot(alpha = 0.6, width = 0.3) +
   labs(
    x = "Aspect",
    y = "Depth (ft)",
    title = "Depth Facet Avalanches by Aspect"
  )




circular_depth<- 
  ggplot(facet_average, aes(x = Aspect, y = count, fill = Aspect)) +
  geom_col(width = 1, color = "black") +
  coord_polar(start = 0) +
  labs(
    title = "Sum of Facet Avalanches by Aspect",
    x = NULL,
    y = "Sum"
  ) +
  theme_minimal() +
  theme(
    plot.background = element_rect(fill = "transparent", color = NA)
  ) +
  theme(
    axis.text.x = element_text(size = 11),
    panel.grid = element_blank()
  )




```

```{r}
#| message: FALSE
#| warning: FALSE

table_facet <- reactable(facet_average %>% 
                     select(Aspect, count, mean_depth, mean_width, mean_vertical, mean_size),
                   theme = reactableTheme(
                     style = list(fontFamily = "sans-serif"),
                     borderColor = "black",
                     backgroundColor = "transparent"
                   ))
```

```{r}
#| message: FALSE
#| warning: FALSE

# size graphs

width_v_depth <- 
  ggplot(utah_avy, aes(x = Depth, y = Width)) +
 geom_density_2d(color = "red") +
  labs(
    x = "Depth (ft)",
    y = "Vertical (ft)",
    title = "Depth vs Vertical Run of Facet Released Avalanches"
  ) +
  theme_minimal()

depth_vert <- 
  ggplot(utah_avy, aes(x = Depth, y = Vertical)) +
 geom_density_2d(color = "blue") +
  labs(
    x = "Depth (ft)",
    y = "Width (ft)",
    title = "Depth vs Width of Facet released Avalanches"
  ) +
  theme_minimal()

```

```{r}
# size

size_facet <- 
  ggplot(facet_average, aes(x = Aspect, y = mean_size, fill = Aspect)) +
  geom_col()

```













# Utah Avalanches

## 

:::{.card title = "Study Area"}

```{r}
region_leaf
```

:::

### 

```{r}
#| content: valuebox
#| title: "Total Avalanches 2020-25"
list(
  icon = 'snow',
  color = "danger",
  value = "4020"
)
```

```{r}
#| content: valuebox
#| title: "Region with Most Avalanches"
list(
  icon = 'water',
  color = "danger",
  value = "Salt Lake"
)
```

```{r}
#| content: valuebox
#| title: "Most Common Weak layer"
list(
  icon = 'grid',
  color = "danger",
  value = "Facets"
)
```

## 

<div>

This dashboard aims to visualize the relationship between facet layer failures and different size variables of avalanches. As seen in the chart above, facet layer failures tend to make up the majority of Avalanches throughout the Utah forecast region. Data obtained through utahavalanchecenter.org.

</div>

:::{.card title = "Avalanches by Weak Layer Failure 2020-25"}

```{r}
year_slide
```

:::

# Facets by Aspect

## 

:::{.card title = "Facet Avalanches by Aspect"}
```{r}
circular_depth
```

```{r}
size_facet
```
:::


##

```{r}
#| content: valuebox
#| title: "Aspect with the most Facet released Avalanches"
list(
  icon = 'compass',
  color = "translucent",
  value = "Northeast"
)
```

```{r}
#| content: valuebox
#| title: "Aspect with largest Avalanches on average"
list(
  icon = 'water',
  color = "translucent",
  value = "South"
)
```



# Facet size 



## Graphs

:::{layout-ncol=2}
:::{.card title = "Size Structure of Facet Avalanches"}
```{r}
width_v_depth
```
:::

:::{.card}
```{r}
depth_vert
```
:::
:::


:::{.card title = "Size of Facet Layers by Variable"}
```{r}
table_facet
```
:::

###


```{r}
#| content: valuebox
#| title: "Average Depth of Facet layer"
list(
  icon = 'rulers',
  color = "purple",
  value = "2.7 ft"
)
```

```{r}
#| content: valuebox
#| title: "Average Width of Avalanche by Facet"
list(
  icon = 'arrow-left-right',
  color = "purple",
  value = "230 ft"
)
```

```{r}
#| content: valuebox
#| title: "Average Vertical Run of Avalanche by Facet"
list(
  icon = "chevron-double-down",
  color = "purple",
  value = "624 ft"
)
```




# Discussion

:::{.card title = "Results"}

-   Overview

The plots shown in this dashboard provide an in depth understanding of where facet avalanches are most likely to occur in relation to aspect, as well as how the depth of the facet layer affects the size of the avalanche.

-   Discussion

I found that while southern aspects had the deepest average depth (3.1 ft), they had the least amount of facet released avalanches out of all aspects, while the majority (384) occurred on North Eastern facing slopes. This can be attributed to the difference in sun that each aspect receives. A southern slope receives the most sun out of any, which in the early season can melt early snowfall, reducing the risk of that early snow rotting until the next snowfall. A more northerly aspect does not receive as much sun, allowing a early snow-pack to sit and rot until normal season conditions persist. 


