---
title: "408_avalanche_dash"
format: dashboard
author: "Sam Pruitt"
editor: visual
orientation: rows
css: stylez.css
number-sections: FALSE
---



```{r setup}
#| message: FALSE

library(knitr)
library(plotly)
library(tidyverse)
library(dataRetrieval)
library(lubridate)
library(sf)
library(ggspatial)
library(prettymapr)
library(lubridate)
library(rosm)
library(bsicons)
library(viridis)
library(shiny)
library(leaflet)
```

```{r}
# read csv
avalanches <- read.csv("avalanches.csv")


# Read zones
zones <- st_read("avalanche_center_forecast_zones.shp", quiet = TRUE)

zones <- st_transform(zones, 4326)


# cleaning names and new df
avalanches <- avalanches |>
  rename(Who = Trigger,
         Trigger = Trigger..additional.info) 

utah_avy <- avalanches |>
  select(-Comments.4, - Comments.3, -Comments.2, - Comments.1, - Weather.Conditions.and.History, -Terrain.Summary, -Accident.and.Rescue.Summary)


  

# creating season DF
utah_avy$Date <- as.Date(utah_avy$Date, format = "%m/%d/%Y")

utah_avy <- utah_avy |>
  filter(Date >= as.Date("2024-10-01") &
         Date <= as.Date("2025-06-01")) 
  


# split into lat and long
utah_avy <- utah_avy |>
  separate(Coordinates, into = c("latitude", "longitude"), sep = ",") |>
  mutate(
    latitude = as.numeric(latitude),
    longitude = as.numeric(longitude)
  )

utah_avy <- utah_avy |>
  filter(!is.na(longitude) & !is.na(latitude))

# convert to spatial points
utah_avy <- st_as_sf(utah_avy,
                       coords = c("longitude", "latitude"),
                       crs = 4326)

# Match Shapefile CRS

utah_avy <- st_transform(utah_avy, st_crs(zones))


# Depth 

utah_avy <- utah_avy |>
  mutate(
    Depth = case_when(
      str_detect(Depth, "'") ~ as.numeric(str_remove(Depth, "'")) * 12,
      str_detect(Depth, '"') ~ as.numeric(str_remove(Depth, '"')),
      TRUE ~ as.numeric(Depth)
    )
  ) 

# Width

utah_avy <- utah_avy |>
  mutate(
    Width = case_when(
      str_detect(Width, "'") ~ as.numeric(str_remove(Width, "'")),
      TRUE ~ as.numeric(Width)
    )
  )

# Vertical

utah_avy <- utah_avy |>
  mutate(
    Vertical = Vertical |>
      str_replace_all(",", "") |>
      str_replace_all("'", "") |>
      as.numeric()
  )

# Aspect

utah_avy$Aspect <- as.factor(utah_avy$Aspect)



```

```{r}
#| message: false
#| echo: false
# map

region_pal <- colorFactor(
  palette = "Set1",
  domain = utah_avy$Region
)


region_leaf <- leaflet() |>
addProviderTiles("Esri.WorldTopoMap") |>
  addPolygons(
    data = zones,
    color = 'red',
    ) |>
  addCircleMarkers(
    data = utah_avy,
    color = ~region_pal(Region),
    fillColor = ~region_pal(Region),
    fillOpacity = .8,
    radius = 6,
    stroke = F
  ) |>
  addLegend(
    "bottomright",
    pal = region_pal,
    values = utah_avy$Region,
    title = "Region"
  )



```

```{r}
# plot showing weak layer and then mean depth of each weak layer

facets <- utah_avy |>
  filter(`Weak.Layer` == "Facets") |>
  select(Width, Depth, Aspect, Vertical)
  
facets <- facets[-115,]
  
facet_plot <-
  ggplot(facets, aes(x = Depth, y = Width, color = Aspect)) +
  geom_point(color = 'black') +
  geom_smooth(method = "loess", se = F) +
  labs(
    x = "Depth",
    y = "Width",
    title = "Size of Facet problem Avalanches, by Aspect"
  )
  


  
```

```{r}
#| warning: FALSE
# Weak Layers


 weak_layers <- ggplot(utah_avy, aes(x = Weak.Layer,na.rm = TRUE, fill = Weak.Layer)) +
  geom_bar() +
  theme(axis.text.x = element_text(angle = 25, hjust = 1)) +
   labs(
     x = "Weak Layer",
     y = "Count",
     title = "Avalanches by Weak Layer"
   )
  
```

```{r}
#| warning: FALSE
# Weak_layer by size



p <- plot_ly(facets,
             x = ~Depth,
             y = ~Width,
             z = ~Vertical,
             color = ~Aspect,
             colors = "Dark2", 
             type = "scatter3d",
             mode = "markers") |>
     layout(
    title = list(
      text = "Size Structures of Facet Problem Avalanches",
      font = list(size = 16)
    ),
    scene = list(
      xaxis = list(
        title = "Depth (in)",
        showgrid = T,
        zeroline = TRUE
      ),
      yaxis = list(
        title = "Width (ft)",
        showgrid = T,
        zeroline = TRUE
      ),
      zaxis = list(
        title = "Vertical (ft)",
        showgrid = T,
        zeroline = TRUE
      ),
      camera = list(eye = list(x = 1.6, y = 1.6, z = 1.2))
    )
  )

```

```{r}
# Facet ggplot

facet_facet <- 
  ggplot(facets, aes(x = Depth, y = Width, color = Aspect)) +
  geom_jitter() +
  geom_smooth(method = "lm", se = F) +
  labs(
    title = "Depth and Width of Facet Avalanches"
  )


facet_aspect <-
  ggplot(facets, aes(x = Aspect, fill = Aspect)) +
  geom_bar() +
   labs(
    x = "Aspect",
    y = "Count",
    title = "Facet Avalanches by Aspect"
  )

  



```

# 2024-25 Avalanches

::: {.card title = "2024-25 Avalanches by Region"}

```{r}
region_leaf
```

:::

## 

<div>

An analysis of different weak layers during the 2024-25 Avalanche Season in Utah, USA. This Dashboard aims to visualize relationships between different weak layer failures and how they impact or are impacted by other variables. Data obtained through utahavalanchecenter.org.

</div>

## 

```{r}
#| content: valuebox
#| title: "Total Avalanches"
list(
  icon = 'snow',
  color = "primary",
  value = "786"
)
```

```{r}
#| content: valuebox
#| title: "Region With the most Avalanche Activity"
list(
  icon = 'water',
  color = "primary",
  value = "Salt Lake"
)
```

```{r}
#| content: valuebox
#| title: "Total Deaths"
list(
  icon = "x",
  color = "primary",
  value = "5"
)
```

# Weak Layers

## 

::: {.card style="background: transparent; border: none;"}
```{r}
weak_layers
```
:::

# Facets

<div>

Avalanches that broke on Facet layers were the most common during the 2024-25 season. The below plots help illustrate how the depth of a facet layer can influence the magnitude of an avalanche.

</div>

## 

<div>

```{r}
p
```

</div>

### 

```{r}
facet_facet
```

```{r}
facet_aspect
```
