---
title: "utah_avy_dash"
format: dashboard
editor: visual
css: stylez.css
orientation: columns
---

```{r setup}
#| message: FALSE

library(knitr)
library(plotly)
library(tidyverse)
library(dataRetrieval)
library(lubridate)
library(sf)
library(ggspatial)
library(prettymapr)
library(lubridate)
library(rosm)
library(bsicons)
library(viridis)
library(shiny)
library(leaflet)
library(reactable)
library(ggridges)
```

```{r}
#| warning: FALSE
#| message: FALSE

# read csv
avalanches <- read.csv("avalanches.csv")


# Read zones
zones <- st_read("avalanche_center_forecast_zones.shp", quiet = TRUE)

zones <- st_transform(zones, 4326)


# cleaning names and new df
avalanches <- avalanches |>
  rename(Who = Trigger,
         Trigger = Trigger..additional.info) 

utah_avy <- avalanches |>
  select(-Comments.4, - Comments.3, -Comments.2, - Comments.1, - Weather.Conditions.and.History, -Terrain.Summary, -Accident.and.Rescue.Summary) |>
  filter(Weak.Layer != "")
  

# creating season DF
utah_avy$Date <- as.Date(utah_avy$Date, format = "%m/%d/%Y")

utah_avy <- utah_avy |>
  filter(Date >= as.Date("2020-01-01") &
         Date <= as.Date("2025-07-01")) |>
  filter(Region %in% c("Abajos", "Logan", "Moab", "Ogden", "Provo", "Salt Lake", "Skyline", "Southwest", "Uintas"))



# split into lat and long
utah_avy <- utah_avy |>
  separate(Coordinates, into = c("latitude", "longitude"), sep = ",") |>
  mutate(
    latitude = as.numeric(latitude),
    longitude = as.numeric(longitude)
  )

utah_avy <- utah_avy |>
  filter(!is.na(longitude) & !is.na(latitude))

# convert to spatial points
utah_avy <- st_as_sf(utah_avy,
                       coords = c("longitude", "latitude"),
                       crs = 4326)

# Match Shapefile CRS

utah_avy <- st_transform(utah_avy, st_crs(zones))


# Depth 

utah_avy <- utah_avy |>
  mutate(
    Depth = case_when(
      str_detect(Depth, "'") ~ as.numeric(str_remove(Depth, "'")) * 12,
      str_detect(Depth, '"') ~ as.numeric(str_remove(Depth, '"')),
      TRUE ~ as.numeric(Depth)
    )
  )  |>
  mutate(
    Depth = Depth / 12
  )

# Width

utah_avy <- utah_avy |>
  mutate(
    Width = case_when(
      str_detect(Width, "'") ~ as.numeric(str_remove(Width, "'")),
      TRUE ~ as.numeric(Width)
    )
  )

# Vertical

utah_avy <- utah_avy |>
  mutate(
    Vertical = Vertical |>
      str_replace_all(",", "") |>
      str_replace_all("'", "") |>
      as.numeric()
  )

# Aspect

utah_avy$Aspect <- as.factor(utah_avy$Aspect)

# Create year column

utah_avy <- utah_avy |>
  mutate(year = format(Date, "%Y"),
         month = month(Date, label = TRUE, abbr = TRUE))


  



```

```{r}
#| message: FALSE
#| warning: FALSE
# Map

region_pal <- colorFactor(
  palette = "Set1",
  domain = zones$name
)


region_leaf <- leaflet() |>
addProviderTiles("Esri.WorldTopoMap") |>
  
  addPolygons(
    data = zones,
    color = ~region_pal(name),
    fillColor =  ~region_pal(name),
    fillOpacity = .6
    ) |>
  addCircleMarkers(
    data = utah_avy,
    color = ~region_pal(Region),
    radius = 6,
    stroke = FALSE,
    fillOpacity = 1,
    group = ~Weak.Layer,
    popup = ~paste0(
  "<table>",
  "<tr><td><b>Date</b></td><td>", Date, "</td></tr>",
  "<tr><td><b>Region</b></td><td>", Region, "</td></tr>",
  "<tr><td><b>Depth</b></td><td>", Depth, "</td></tr>",
  "<tr><td><b>Width</b></td><td>", Width, "</td></tr>",
  "<tr><td><b>Vertical</b></td><td>", Vertical, "</td></tr>",
  "<tr><td><b>Weak Layer</b></td><td>", Weak.Layer, "</td></tr>",
  "</table>"
) )|>
  addLayersControl(
    overlayGroups = unique(utah_avy$Weak.Layer),
    options = layersControlOptions(collapsed = TRUE)
  ) |>
  addLegend(
    "bottomright",
    pal = region_pal,
    values = utah_avy$Region,
    title = "Region"
  )

```

```{r}
#| message: FALSE
#| warning: FALSE

avy_counts <- utah_avy |>
  st_drop_geometry() |>
  filter(
    !is.na(year),
    !is.na(Region), Region != "",
    !is.na(Weak.Layer), Weak.Layer != ""
  ) |>
  count(year, Region, Weak.Layer)

year_slide <- plot_ly(
  data = avy_counts,
  x = ~Region,
  y = ~n,
  frame = ~year,
  type = "bar",
  color = ~Weak.Layer
) |>
  layout(barmode = "stack") 
```

```{r}
#| message: FALSE
#| warning: FALSE
#| fig-format: png
#| fig-bg: transparent

# Make Facet DF
facets <- utah_avy |>
  filter(`Weak.Layer` == "Facets") |>
  select(Width, Depth, Aspect, Vertical, year, month) |>
  filter(
    !is.na(year),
    !is.na(Aspect), Aspect != "",
    !is.na(Width), Width != "",
    !is.na(Depth), Depth != "",
    !is.na(month), month != "",
  )

# Mean facet depth by aspect
facet_average <- facets |>
  group_by(Aspect) |>
  summarise(
    count = n(),
    mean_depth = mean(Depth, na.rm = TRUE),
    mean_width = mean(Width, na.rm = TRUE), 
    mean_vertical = mean(Vertical, na.rm = TRUE)
  ) |>
  mutate(
    mean_depth = round(mean_depth, 1),
    mean_width = round(mean_width, 1),
    mean_vertical = round(mean_vertical, 1)
  )
 
```

```{r}
#| fig-height: 10
#| fig-width: 8
facet_aspect <-
  ggplot(facets, aes( x = "", y = Depth, color = Aspect)) +
  geom_jitter(size = 3) +
  geom_violin(alpha = 0.6, width = 0.6)+
  geom_boxplot(alpha = 0.6, width = 0.3) +
   labs(
    x = NULL,
    y = "Depth (ft)",
    title = "Depth Facet Avalanches by Aspect"
  ) +
  facet_wrap(facets$Aspect, ncol = 4, scales = "free_y") +
  theme(
    legend.position = "none"
  )

```

```{r}
#| message: FALSE
#| warning: FALSE

table_facet <- reactable(facet_average %>% 
                     select(Aspect, count, mean_depth, mean_width, mean_vertical),
                   theme = reactableTheme(
                     style = list(fontFamily = "sans-serif"),
                     borderColor = "black",
                     backgroundColor = "translucent"
                   ))
```

```{r}
#| message: FALSE
#| warning: FALSE

# size graphs

width_v_depth <- 
  ggplot(utah_avy, aes(x = Depth, y = Width)) +
 geom_density_2d(color = "red") +
  labs(
    x = "Depth (ft)",
    y = "Vertical (ft)",
    title = "Depth vs Vertical Run of Facet Released Avalanches"
  ) +
  theme_minimal()

depth_vert <- 
  ggplot(utah_avy, aes(x = Depth, y = Vertical)) +
 geom_density_2d(color = "blue") +
  labs(
    x = "Depth (ft)",
    y = "Width (ft)",
    title = "Depth vs Width of Facet released Avalanches"
  ) +
  theme_minimal()

```






# Utah Avalanches

## Map

:::{.card title = "Study Area"}
```{r}
region_leaf
```
:::

### valuebox

```{r}
#| content: valuebox
#| title: "Total Avalanches 2020-25"
list(
  icon = 'snow',
  color = "danger",
  value = "4020"
)
```

```{r}
#| content: valuebox
#| title: "Region with Most Avalanches"
list(
  icon = 'water',
  color = "danger",
  value = "Salt Lake"
)
```

```{r}
#| content: valuebox
#| title: "Most Common Weak layer"
list(
  icon = 'grid',
  color = "danger",
  value = "Facets"
)
```

## facts

This dashboard aims to visualize different variables in facet layer avalanches. As seen in the map and plot below, avalanches by facet layer are the most common avalanche problem throughout the Utah forecast region. Data obtained through utahavalanchecenter.org.



:::{.card title = "Avalanches by Weak Layer Failure 2020-25"}
```{r}
year_slide
```
:::

# Facets by Aspect

### aspect

:::{.card title = "Depth Facet Avalanche by Aspect" }
```{r}
#| fig-width: 20
#| fig-height: 10
facet_aspect
```
:::


:::{.card title = "Size Variables by Aspect" }
```{r}
table_facet
```
:::



# Facet size 



## Graphs


:::{layout-ncol =2}
```{r}
width_v_depth
```

```{r}
depth_vert
```
:::



### Valueboxes


```{r}
#| content: valuebox
#| title: "Mean Depth of Facet layer"
list(
  icon = 'rulers',
  color = "purple",
  value = "2.7 ft"
)
```

```{r}
#| content: valuebox
#| title: "Mean Width of Avalanche by Facet"
list(
  icon = 'arrow-left-right',
  color = "purple",
  value = "230 ft"
)
```

```{r}
#| content: valuebox
#| title: "Mean Vertical Run of Avalanche by Facet"
list(
  icon = "chevron-double-down",
  color = "purple",
  value = "624 ft"
)
```




# Discussion

:::{.card title = "Results"}

- Discussion

While southern aspects yielded the highest average facet layer depth (3.1 ft), they had they lowest frequency of facet related avalanches. In contrast, the most avalanches (n = 384) occurred on northeastern aspects. This shows that there is a relationship between aspect and avalanche occurrence.

This pattern across aspects is most likely driven by the variation in solar radiation. South facing slopes receive much more sun, especially in the eaely season. This promotes freeze-thaw cycles, or complete melting of early season snow, which can prohibit persistent weak layers (PWL's) such as facets to form. As a result, while facets may be deeper in the snow pack later on, the structure is less likely to have widespread facet layers. 

In contrast, North and Northeast facing slopes receive much less solar radiation than that of south facing slopes. This creates a much colder atmosphere, and when the temperature of deeper snow is warmer than the surface snow, can cause water vapor to move upward. This vapor recrystallizes as angular snow, forming a very weak and sugary layer. These layers cannot support high volumes of new snow, causing it to release as an avalanche. 

The mean facet avalanche depth across all observations was 2.7 ft, with an average width of 230 ft and a mean vertical run of 624 ft. These means suggest that even with relatively shallow facets, large avalanches are common. This is consistent with Persistent Weak Layer behavior.

Observations were found in Utah, where a colder and drier climate exists. Compared to a location such as Washington, with a maritime climate, facet observations may differ.

Links for future implications :

- https://utahavalanchecenter.org/

- https://avtraining.org/

:::


